Human Factors Influencing Visual Statistical Inference

### 2013-06-09 sociology_paper.tex
some allignment in the table
factor effects updated

There have been \dc{ten} experiments conducted using Amazon's Mechanical Turk~\cite{turk} that are being used to evaluate the the effect of the observer. Table \ref{tbl:visual_stat} summarizes these. All of these experiments are used to evaluate the effects of demographic factors. Experiments 5,6 and 7 are used to study learning trend of the observer. The design of experiment 9 enabled effect of position in the lineup, and the sample of nulls, to be evaluated. Section \ref{sec:factor_performance} describes the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the methods used to assess the effects, and Section \ref{sec:result_socio} describes the results.

@@

The visual test statistic is chosen so that it displays a specific patter in case the null hypothesis not true. Thus the most important factor that help an observer to correctly evaluate a lineup is the presence of any detectable signal in the data. On the other hand, if the null hypothesis is not true, visual test statistic should not display any distinguishable pattern. In fact, some of the null plots may appear to be the most different plot in a lineup influencing the observer to chose a plot different than actual data plot. This is an elegant feature of the lineup. It force the observer to chose a wrong plot when the null hypothesis should not be rejected.

@@

@@

These questions are very crucial for the power of visual test. They help observer think in a controlled direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These question help observer not to be misguided by those patterns. To review further on this \citet{majumder:2013} have also asked why the observers choose a specific plot.

@@

\subsection{Demographics of Observer}  Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. Also, to meet the Institutional Regulatory Board (IRB) requirement is is necessary to exclude any subject less than 18 years of old to participate human subject experiment. Thus we intend to only focus on the performance of observer with age 18 years or older. But still there may be variation in different age groups. Recruiting subjects with a well variety of age groups may be necessary to have control over this variability.

@@

\subsection{Learning Trend of the Observer} When an observer evaluates a lineup he or she may learn something from the experience. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it could influence the performance of the observer in some way. This paper investigates the learning trend with multiple simulation experiments.

@@

We design and develop a web site which enables to display the lineups to the observers as per experimental design. The MTurk workers are redirect to the web site and the data are collected, stored automatically into a local database server. Demographic informations such as age group, gender and education levels were collected. The time taken for each evaluation is computed based on the time the plot was shown and the time the feedback was received. The location of the observer is determined by the ip address of the observer.

@@

\cline{2-5} \cline{7-10} \cline{12-15} 
 Effect& Est & SE & Zval & $p$-value && Est & SE & Zval & $p$-value && Est & SE & Zval & $p$-value \\ 
  \hline
Fixed &  &  &  & && &  & & & & & &  & \\ 
$\mu$ & 3.817 & 0.039 & 97.38 & $<$0.001 &   & 3.901 & 0.033 & 118.19 & $<$0.001 &   & 3.731 & 0.054 & 69.04 & $<$0.001 \\ 
  $\alpha_1$ & 0.326 & 0.035 & 9.35 & $<$0.001 &   & 0.335 & 0.029 & 11.4 & $<$0.001 &   & 0.280 & 0.050 & 5.63 & $<$0.001 \\ 
  $\alpha$ & -0.038 & 0.004 & -9.3 & $<$0.001 &   & -0.039 & 0.004 & -10.19 & $<$0.001 &   & -0.029 & 0.007 & -4.22 & $<$0.001 \\ 
\multicolumn{2}{l}{Random}   &  &  &  &  & & &&& & & &  & \\ 
  $\sigma^2_a$ & 0.001 & 0.027 &  &  &   & 0.002 & 0.045 &  &  &   & 0.002 & 0.049 &  &  \\ 
  $\sigma^2_u$ & 0.259 & 0.509 &  &  &   & 0.245 & 0.495 &  &  &   & 0.134 & 0.366 &  &  \\ 
  $\sigma^2_l$ & 0.008 & 0.091 &  &  &   & 0.040 & 0.199 &  &  &   & 0.055 & 0.235 &  &  \\ 
  $\sigma^2$ & 0.211 & 0.460 &  &  &   & 0.251 & 0.501 &  &  &   & 0.206 & 0.454 &  &  \\ 
   \hline
\end{tabular}
}
\label{tbl:model_result_time}
\end{table}

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Model \eqref{eqn:trend_response} fitted for log time taken to evaluate a lineup. None of the fixed factor effects of attempt (2-10) seem to be significantly different from each other at \%5 level in all three experiments 5, 6 and 7.}
\scalebox{0.9}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

>>>>

There have been ten experiments conducted using Amazon's Mechanical Turk~\cite{turk} that are being used to evaluate the the effect of the observer. Table \ref{tbl:visual_stat} summarizes these. All of these experiments are used to evaluate the effects of demographic factors. Experiments 5,6 and 7 are used to study learning trend of the observer. The design of experiment 9 enabled effect of position in the lineup, and the sample of nulls, to be evaluated. Section \ref{sec:factor_performance} describes the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the methods used to assess the effects, and Section \ref{sec:result_socio} describes the results.

@@

The visual test statistic is chosen so that it displays a specific pattern in case the null hypothesis is not true. Thus the most important factor that help an observer to correctly evaluate a lineup is the presence of any detectable signal in the data. On the other hand, if the null hypothesis is not true, visual test statistic should not display any distinguishable pattern. In fact, some of the null plots may appear to be the most different plot in a lineup influencing the observer to chose a plot different than actual data plot. This is an elegant feature of the lineup. It force the observer to chose a wrong plot when the null hypothesis should not be rejected.

@@

Table \ref{tbl:visual_stat} shows the visual test statistics used in this study. For both experiment 2 and 3 a scatterplot is used but a regression line fitted through the points is overlaid for experiment 2. It is easier to spot the slope of the line compared to just the scatterplot itself. But it appears that people are better in noticing the unusual pattern in the data which is some contamination purposefully added for experiment 3. It is observed that for experiment 3 the performance of the observers are much better compared to experiment 2 \citep{majumder:2013}.

@@

These questions are very crucial for the power of visual test. They help observer think in a meaningful direction. Notice that there may be unnecessary patterns in the actual data plot which may not necessarily indicate the existence of the significant signal in the plot. These question help observer not to be misguided by those patterns. To review further on this \citet{majumder:2013} have also asked why the observers choose a specific plot.

@@

\subsection{Demographics of Observer}  Some of the demographics of the observer such as age, gender, education level and geographical location may have effect on how an observer would examine the lineup. This may produce some variability in the performance of the observer. For example, a high school student may not respond same as a well educated person while evaluating a lineup. There may be variation in different age groups as well. Recruiting subjects with a well variety of age groups is be necessary to study this variability.

To meet the Institutional Regulatory Board (IRB) requirement, it is necessary to exclude any subject less than 18 years of old to participate human subject experiment. Thus we intend to only focus on the performance of observer with age 18 years or older. We don't expect many older people to participate our study and we are interested of age from 18 to 65 years.

@@

We don't expect much difference in performances between male and female observers. Gender may have some influence through the education level and geographical locations since some location may have small number of educated female population. We intend to include similar number of male and female subjects in our study so that this can be examined.

\subsection{Learning Trend of the Observer} When an observer evaluates a lineup he or she may learn something from the experiences of the earlier observations. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it could influence the performance of the observer in some way.

Learning may occur in two different ways. An observer can become more trained on the lineup structure and the pattern in the plots shown. This could be learning in proportion of correct evaluation. Or, he or she may become efficient in responding faster in the later evaluations. This could be observed as the time taken for evaluating a lineup. This paper investigates both of these learning trends with multiple simulation experiments.

When an observer evaluate a lineup for the first time, he or she has to read through many instructions and become familiar with the environment of the experiment. We expect that this would require much more time compared to the rest of the evaluations. Time taken for the second or successive attempt would be much lower than the time taken for the first lineup. We attribute this to the adaptation to the experiment environment, not a learning of evaluating a lineup.

@@

We designed and developed a web site which enables to display the lineups to the observers as per experimental design. The MTurk workers were redirected to the web site and the data were collected, stored automatically into a local database server. Demographic informations such as age group, gender and education levels were also collected. The time taken for each evaluation is computed based on the time the plot was shown and the time the feedback was received. The location of the observer is determined by the ip address of the observer.

@@

\cline{2-5} \cline{7-10} \cline{12-15} 
 Effect& Est & SE & Zval & $p$-value && Est & SE & Zval & $p$-value && Est & SE & Zval & $p$-value \\ 
  \hline
Fixed &  &  &  & && &  & & & & & &  & \\ 
$\mu$ & 3.817 & 0.039 & 97.38 & $<$0.001 &   & 3.901 & 0.033 & 118.19 & $<$0.001 &   & 3.731 & 0.054 & 69.04 & $<$0.001 \\ 
  $\alpha_1$ & 0.326 & 0.035 & 9.35 & $<$0.001 &   & 0.335 & 0.029 & 11.40 & $<$0.001 &   & 0.280 & 0.050 & 5.63 & $<$0.001 \\ 
  $\alpha$ & -0.038 & 0.004 & -9.30 & $<$0.001 &   & -0.039 & 0.004 & -10.19 & $<$0.001 &   & -0.029 & 0.007 & -4.22 & $<$0.001 \\ 
\multicolumn{2}{l}{Random}   &  &  &  &  & & &&& & & &  & \\ 
  $\sigma^2_a$ & 0.001 & 0.027 &  &  &   & 0.002 & 0.045 &  &  &   & 0.002 & 0.049 &  &  \\ 
  $\sigma^2_u$ & 0.259 & 0.509 &  &  &   & 0.245 & 0.495 &  &  &   & 0.134 & 0.366 &  &  \\ 
  $\sigma^2_l$ & 0.008 & 0.091 &  &  &   & 0.040 & 0.199 &  &  &   & 0.055 & 0.235 &  &  \\ 
  $\sigma^2$ & 0.211 & 0.460 &  &  &   & 0.251 & 0.501 &  &  &   & 0.206 & 0.454 &  &  \\ 
   \hline
\end{tabular}
}
\label{tbl:model_result_time}
\end{table}

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Model \eqref{eqn:trend_response} fitted for log time taken to evaluate a lineup. None of the fixed factor effects of attempt (2-10) seem to be significantly different from each other at \%5 level in all three experiments 5, 6 and 7. For experiment 7 subject specific variation is very small on the other hand lineup variance is much higher compared to the other two experiments.}
\scalebox{0.9}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\
